---
title: "Lab 11 - Interactive Visualization"
output: html_document
---


# Learning Goals

- Read in and process the COVID dataset from the New York Times GitHub repository
- Create interactive graphs of different types using `plot_ly()` and `ggplotly()` functions
- Customize the hoverinfo and other plot features
- Create a Choropleth map using `plot_geo()`

# Lab Description

We will work with COVID data downloaded from the New York Times. The dataset consists of COVID-19 cases and deaths in each US state during the course of the COVID epidemic.

**The objective of this lab is to explore relationships between cases, deaths, and population sizes of US states, and plot data to demonstrate this**

# Steps

## I. Reading and processing the New York Times (NYT) state-level COVID-19 data

### 0. Install and load libraries

```{r message=FALSE, echo=FALSE, warning=FALSE}

library(data.table)
library(tidyverse)
library(plotly)
library(knitr)
library(widgetframe)
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache=FALSE,
  include=TRUE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px")
```
### 1. Read in the data

- Read in the COVID data with data.table:fread() from the NYT GitHub repository: "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv"
- Read in the state population data with data.table:fread() from the repository: "https://raw.githubusercontent.com/COVID19Tracking/associated-data/master/us_census_data/us_census_2018_population_estimates_states.csv""
- Merge datasets

```{r}

cv_states_readin <- as.data.frame(data.table::fread("https://raw.githubusercontent.com/jeffreyz374/JSC370-Labs-And-Assignments/main/data/covid-states.csv") )


state_pops <- as.data.frame(data.table::fread("https://raw.githubusercontent.com/jeffreyz374/JSC370-Labs-And-Assignments/main/data/us-census-data.csv"))

state_pops$abb <- state_pops$state
state_pops$state <- state_pops$state_name
state_pops$state_name <- NULL

cv_states <- merge(cv_states_readin, state_pops, by="state")
```

### 2. Look at the data

- Inspect the dimensions, `head`, and `tail` of the data
- Inspect the structure of each variables. Are they in the correct format?

```{r}
dim(cv_states)
head(cv_states)
tail(cv_states)
str(cv_states)
```

### 3. Format the data

- Make date into a date variable
- Make state into a factor variable
- Order the data first by state, second by date
- Confirm the variables are now correctly formatted
- Inspect the range values for each variable. What is the date range? The range of cases and deaths?

```{r}

cv_states$date <- as.Date(cv_states$date, format="%Y-%m-%d")


state_list <- unique(cv_states$state)
cv_states$state <- factor(cv_states$state, levels = state_list)
abb_list <- unique(cv_states$abb)
cv_states$abb <- factor(cv_states$abb, levels = abb_list)


cv_states = cv_states[order(cv_states$state, cv_states$date),]


str(cv_states)
head(cv_states)
tail(cv_states)


head(cv_states)
summary(cv_states)

# The earliest date:
min(cv_states$date)

# The latest date:
max(cv_states$date)

# The lowest case count:
min(cv_states$cases)

# The highest case count:
max(cv_states$cases)

# The lowest death count:
min(cv_states$deaths)

# The highest death count:
max(cv_states$deaths)
```

### 4. Add `new_cases` and `new_deaths` and correct outliers

- Add variables for new cases, `new_cases`, and new deaths, `new_deaths`: 
  - Hint: You can set `new_cases` equal to the difference between cases on date i and date i-1, starting on date i=2
  
- Filter to dates after July 1, 2021


```{r}

for (i in 1:length(state_list)) {
  cv_subset = subset(cv_states, state == state_list[i])
  cv_subset = cv_subset[order(cv_subset$date),]

  # add starting level for new cases and deaths
  cv_subset$new_cases = cv_subset$cases[1]
  cv_subset$new_deaths = cv_subset$deaths[1]


  for (j in 2:nrow(cv_subset)) {
    cv_subset$new_cases[j] = cv_subset$cases[j] - cv_subset$cases[j-1] 
    cv_subset$new_deaths[j] = cv_subset$deaths[j] - cv_subset$deaths[j-1]
  }
  
  cv_states$new_cases[cv_states$state==state_list[i]] = cv_subset$new_cases
  cv_states$new_deaths[cv_states$state==state_list[i]] = cv_subset$new_deaths
}

# Focus on recent dates
cv_states <- cv_states %>% dplyr::filter(date >= "2021-07-01")
```

- Use `ggplotly` for EDA: See if there are outliers or values that don't make sense for `new_cases` and `new_deaths`. Which states and which dates have strange values?

```{r}
p1 <- ggplot(cv_states, aes(x = date, y = new_cases, colour = state)) +
  geom_line()

ggplotly(p1)
```

```{r}
p2 <- ggplot(cv_states, aes(x = date, y = new_deaths, colour = state)) +
  geom_line()

ggplotly(p2)
```

- Correct outliers: Set negative values for `new_cases` or `new_deaths` to 0

- Inspect data again interactively

```{r}
cv_states$new_cases[cv_states$new_cases < 0] = 0

p3a <- ggplot(cv_states, aes(x = date, y = new_cases, colour = state)) +
  geom_line() +
  geom_point(size = 0.5, alpha = 0.5)

ggplotly(p3a)

```

```{r}
cv_states$new_deaths[cv_states$new_deaths < 0] = 0

p3b <- ggplot(cv_states, aes(x = date, y = new_deaths, colour = state)) +
  geom_line() +
  geom_point(size = 0.5, alpha = 0.5)

ggplotly(p3b)
```


### 5. Add additional variables

- Add population-normalized (by 100,000) variables for each variable type (rounded to 1 decimal place). Make sure the variables you calculate are in the correct format (`numeric`). You can use the following variable names:
  - `per100k` = cases per 100,000 population
  - `newper100k`= new cases per 100,000
  - `deathsper100k` = deaths per 100,000
  - `newdeathsper100k` = new deaths per 100,000

- Add a "naive CFR" variable representing `deaths / cases` on each date for each state

- Create a dataframe representing values on the most recent date, `cv_states_today`

```{r}

cv_states$per100k =  as.numeric(format(round(cv_states$cases/(cv_states$population/100000),1),nsmall=1))
cv_states$newper100k =  as.numeric(format(round(cv_states$new_cases/(cv_states$population/100000),1),nsmall=1))
cv_states$deathsper100k =  as.numeric(format(round(cv_states$deaths/(cv_states$population/100000),1),nsmall=1))
cv_states$newdeathsper100k =  as.numeric(format(round(cv_states$new_deaths/(cv_states$population/100000),1),nsmall=1))

cv_states = cv_states %>% mutate(naive_CFR = round((deaths*100/cases),2))

max_date <- max(cv_states$date)
cv_states_today = cv_states %>% filter(date==as.Date(max_date))

```

## II. Scatterplots

### 6. Explore scatterplots using `plot_ly()`

- Create a scatterplot using `plot_ly()` representing `pop_density` vs. various variables (e.g. `cases`, `per100k`, `deaths`, `deathsper100k`) for each state on most recent date (`cv_states_today`)
  - Color points by state and size points by state population
  - Use hover to identify any outliers. 


```{r, warning = FALSE, message = FALSE}
cv_states_today %>%
  plot_ly(x = ~pop_density, y = ~cases, type = 'scatter', mode = "markers", 
          color = ~state, size = ~population, sizes = c(5, 70), 
          marker = list(sizemode = 'diameter', opacity = 0.5))
```

- Remove those outliers and replot.

```{r, warning = FALSE, message = FALSE}
cv_states_today %>%
  filter(state != "District of Columbia") %>%
  plot_ly(x = ~pop_density, y = ~cases, color = ~state, type = "scatter", mode = "markers", 
          size = ~population, sizes = c(5, 70), 
          marker = list(sizemode = 'diameter', opacity = 0.5))
```

- Choose one plot. For this plot:
- Add hoverinfo specifying the state name, cases per 100k, and deaths per 100k, similarly to how we did this in the lecture notes
- Add layout information to title the chart and the axes
- Enable `hovermode = "compare"`
  
```{r, warning = FALSE, message = FALSE}
cv_states_today %>%
  filter(state != "District of Columbia") %>%
  na.omit(state) %>%
  plot_ly(x = ~pop_density, y = ~cases, type = "scatter", mode = "markers", 
          color = ~state, size = ~population, sizes = c(5, 70), 
          marker = list(sizemode = 'diameter', opacity = 0.5),
          hover_info = "text", 
          text = ~paste(paste0("State:", state),
                        paste0("Cases per 100k:", per100k),
                        paste0("Deaths per 100k:", deathsper100k),
                        sep = "<br>")
  ) %>%
  layout(title = "Population-normalized cases per 100k",
         yaxis = list(title = "cases per 100k"),
         xaxis = list(title = "population density"),
         hovermode = "compare")
```
  
### 7. Explore scatterplot trend interactively using `ggplotly()`

- For `pop_density` vs. `newdeathsper100k` create a chart with the same variables using `gglotly()`
- Explore the pattern between $x$ and $y$
- Explain what you see. Do you think `pop_density` correlates with `newdeathsper100k`?

```{r}
p4 <- cv_states_today %>%
  filter(state != "District of Columbia") %>%
  ggplot(aes(x = pop_density, y = deathsper100k, colour = state, size = population)) +
  geom_point()

ggplotly(p4)
```

It appears as though the plot depicts that there is a somewhat positive correlation between population density and new deaths per 100,000. This doesn't seem too unusual since it would make sense for states with greater population densities to experience greater spread of the virus and thus incur a higher death rate.

### 8. Multiple line chart

- Create a line chart of the `naive_CFR` for all states over time using `plot_ly()`
  - Use the zoom and pan tools to inspect the `naive_CFR` for the states that had an increase in September. How have they changed over time?
- Create one more line chart, for Florida only, which shows `new_cases` and `new_deaths` together in one plot. Hint: use `add_layer()`
  - Use hoverinfo to "eyeball" the approximate peak of deaths and peak of cases. What is the time delay between the peak of cases and the peak of deaths?
  

```{r, warning = FALSE, message = FALSE}
cv_states %>%
  plot_ly(x = ~date, y = ~naive_CFR, color = ~state, type = "scatter", mode = "lines")
```

```{r}
cv_states %>%
  filter(state == "Florida") %>%
  plot_ly(x = ~date, y = ~new_cases, type = "scatter", mode = "lines", name = "cases") %>%
  add_lines(x = ~date, y = ~new_deaths, type = "scatter", mode = "lines", name = "deaths")
```

By looking at and hovering over the charts above, we can see that the approximate delay between spikes in both new cases and new deaths is 1 week, or 7 days. This makes sense since the incubation period of the virus is around 7-14 days, so new cases and new deaths will follow a similar pattern of spiking around that same timeframe.

### 9. Heatmaps

Create a heatmap to visualize `new_cases` for each state on each date greater than July 1st, 2021
- Start by mapping selected features in the dataframe into a matrix using the **tidyr** package function `pivot_wider()`, naming the rows and columns, as done in the lecture notes
- Use `plot_ly()` to create a heatmap out of this matrix. Which states stand out?


```{r}
cv_states_mat <- cv_states %>%
  select(state, date, new_cases) %>%
  filter(date > "2021-07-01")

cv_states_mat2 <- as.data.frame(pivot_wider(cv_states_mat, 
                                            names_from = state, 
                                            values_from = new_cases))

cv_states_mat2 <- cv_states_mat2 %>%
  column_to_rownames("date") %>%
  as.matrix()

plot_ly(x = colnames(cv_states_mat2), 
        y = rownames(cv_states_mat2), 
        z = ~cv_states_mat2,
        type = "heatmap")

```

In the heatmap above, the state of California stands out as the only state that had a time period where a significant spike in new cases (likely fueled by the Omicron variant) occurred.

# Deliverables 

Lab 10b questions 1-2, lab 11 questions 0-10. Upload html or pdf for both lab Rmd's to quercus. 
