---
title: "Lab 04 - Data Visualization"
output:
  html_document:
    html_preview: no
  pdf_document: default
link-citations: yes
---

```{r setup, echo = FALSE, include=FALSE, eval = TRUE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(data.table)
library(leaflet)
library(tidyverse)
library(gifski)
library(av)
library(gganimate)
```

# Learning Goals

-   Read in and prepare the meteorological dataset
-   Create several graphs with different `geoms()` in `ggplot2`
-   Create a facet graph
-   Conduct some customizations of the graphs
-   Create a more detailed map using `leaflet()`

# Lab Description

We will again work with the meteorological data presented in lecture.

**The objective of the lab is to examine the association between weekly average dew point temperature and wind speed in four regions of the US and by elevation.**

# Steps

### 1. Read in the data

First download and then read in with data.table:fread()

This is the same met_data as we used in lab 03, so you can choose to move that file to your current project directory instead of re-downloading.

```{r, echo=FALSE, message=FALSE}
met <- data.table::fread("https://raw.githubusercontent.com/jeffreyz374/JSC370-Labs-And-Assignments/main/data/met_all.gz")
```

### 2. Prepare the data

-   Remove temperatures less than -17C
-   Make sure there are no missing data in the key variables coded as 9999, 999, etc
-   Generate a date variable using the functions `as.Date()` (hint: You will need the following to create a date `paste(year, month, day, sep = "-")`).
-   Using the `data.table::week` function, keep the observations of the first week of the month.
-   Compute the mean by station of the variables `temp`, `rh`, `wind.sp`, `vis.dist`, `dew.point`, `lat`,`lon`, and `elev`.
-   Create a region variable for NW, SW, NE, SE based on lon = -98.00 and lat = 39.71 degrees
-   Create a categorical variable for elevation (breakpoint 252 m)

```{r, eval = TRUE, echo = FALSE}
met %>%
  filter(temp >= -17) %>%
  na_if(9999) %>%
  mutate(data = as.Date(paste(year, month, day, sep = "-"))) %>%
  filter(day <= 7) %>%
  group_by(USAFID) %>%
  summarise(temp = mean(temp, na.rm = TRUE),
            rh = mean(rh, na.rm = TRUE),
            wind.sp = mean(wind.sp, na.rm = TRUE),
            vis.dist = mean(vis.dist, na.rm = TRUE), 
            dew.point = mean(dew.point, na.rm = TRUE),
            lat = mean(lat),
            lon = mean(lon),
            elev = mean(elev, na.rm = TRUE)) %>%
  mutate(region = ifelse(lat > 39.71, "north", "south")) %>%
  mutate(region = paste(region, ifelse(lon > -98, "east", "west"))) %>%
  mutate(elev_cat = ifelse(elev > 252, "high", "low")) -> met_avg
```

### 3. Use `geom_violin` to examine the wind speed and dew point temperature by region

You saw how to use `geom_boxplot` in class. Try using `geom_violin` instead (take a look at the help). (hint: You will need to set the `x` aesthetic to 1)

-   Use facets
-   Make sure to deal with `NA` category
-   Describe what you observe in the graph

```{r, eval = TRUE, echo = FALSE, message = FALSE}
met_avg %>%
  filter(!(region %in% NA)) %>%
  ggplot() +
  aes(x = 1, y = wind.sp, fill = region) +
  geom_violin() +
  facet_wrap(~region)

met_avg %>%
  filter(!(region %in% NA)) %>%
  ggplot() +
  aes(x = 1, y = wind.sp, fill = region) +
  geom_boxplot() +
  facet_wrap(~region)

met_avg %>%
  filter(!(region %in% NA)) %>%
  ggplot() +
  aes(x = 1, y = dew.point, fill = region) +
  geom_violin() +
  facet_wrap(~region)

met_avg %>%
  filter(!(region %in% NA)) %>%
  ggplot() +
  aes(x = 1, y = dew.point, fill = region) +
  geom_boxplot() +
  facet_wrap(~region)
```

In the first two plots, we can see that the distributions for average windspeed for the northwest and southwest regions of the United States tend to have a more symmetrical distribution while the corresponding distributions for the northeast and southeast regions tend to be more skewed to the right, especially with some extreme outliers for the northeast. This suggests that the east coast of the United States tends to have more variability in average windspeed in the first week of August than the west coast, which makes sense considering that August is the middle of the annual hurricane season that usually pummels the southeast.

In the next two plots, we can see that the distributions for average dew point temperature for the northeast and southeast look more or less symmetrical with a few outliers, while the corresponding distributions for the northwest and southwest look a little more strange, where the distribution for the southwest looks like it is more skewed left while the distribution for the northwest looks skewed left but also bimodal with peaks at around 8 and 17. This suggests the reverse of the trend concerning windspeed in that the average dew point temperature of the west coast of the United States has more variability in the first week of August than the east coast.

### 4. Use `geom_jitter` with `stat_smooth` to examine the association between dew point temperature and wind speed by region

-   Color points by region
-   Make sure to deal with `NA` category
-   Fit a linear regression line by region
-   Describe what you observe in the graph

```{r, eval = TRUE, echo = FALSE}
met_avg %>%
  filter(!(region %in% NA)) %>%
  ggplot() +
  aes(x = dew.point, y = wind.sp, color = region) +
  geom_jitter() +
  stat_smooth(method = lm)
```

Eyeballing the plot above, it seems like there isn't much association between dew point temperature and windspeed for any of the 4 regions. To investigate this, we can compute the actual correlation coefficients for all 4 of these associations, in order of north east, north west, south east, and south west:
```{r, eval=TRUE, echo=FALSE}
cor(met_avg[met_avg$region == "north east",]$dew.point, met_avg[met_avg$region == "north east",]$wind.sp, use = "complete.obs")

cor(met_avg[met_avg$region == "north west",]$dew.point, met_avg[met_avg$region == "north west",]$wind.sp, use = "complete.obs")

cor(met_avg[met_avg$region == "south east",]$dew.point, met_avg[met_avg$region == "south east",]$wind.sp, use = "complete.obs")

cor(met_avg[met_avg$region == "south west",]$dew.point, met_avg[met_avg$region == "south west",]$wind.sp, use = "complete.obs")
```

With these correlation coefficients, we can confirm that there is very little association between dew point temperature and windspeed in any of the 4 regions.

### 5. Use `geom_bar` to create barplots of the weather stations by elevation category coloured by region

-   Bars by elevation category using `position="dodge"`
-   Change colors from the default. Color by region using `scale_fill_brewer` see [this](http://rstudio-pubs-static.s3.amazonaws.com/5312_98fc1aba2d5740dd849a5ab797cc2c8d.html)
-   Create nice labels on axes and add a title
-   Try a different theme
-   Describe what you observe in the graph
-   Make sure to deal with `NA`

```{r warning=FALSE, message=FALSE, eval = TRUE, echo = FALSE}
met_avg %>%
  filter(!(region %in% NA)) %>%
  ggplot() +
  aes(x = elev_cat, fill = region) +
  geom_bar(position = "dodge") +
  theme_minimal() +
  labs(title = "Number of Weather Stations By Elevation Category and Region", 
       x = "Elevation", y = "Count")
```

First, the bar plot on the left that visualizes the number of weather stations we classified as "high" elevation (i.e. those that more than 252 metres above sea level) shows that the northeast has the greatest number of "high" weather stations, followed by the southwest, southeast, and the northwest. This is a little surprising as a reasonable assumption that could've been made was that the southwest or the northwest would have the tallest bar in this chart as they encompass the Rocky Mountains and the Cascades, respectively. 

Next, the bar plot on the right that visualizes the number of weather stations we classified as "low" elevation shows that the southeast has the greatest number of "low" weather stations by far, followed by the northeast, southwest, and northwest. This plot is a little less surprising, as it would be expected that the flatter southeast would have the most weather stations at a lower elevation.

### 6. Use `stat_summary` to examine mean dew point and wind speed by region with standard deviation error bars

-   Make sure to remove `NA`
-   Use fun.data="mean_sdl" in `stat_summary`
-   Add another layer of `stats_summary` but change the geom to `"errorbar"` (see the help).
-   Describe the graph and what you observe

```{r, eval = TRUE, echo = FALSE}
met_avg %>%
  filter(!(region %in% NA)) %>%
  ggplot() +
  aes(x = region, y = wind.sp) +
  stat_summary(fun.data = "mean_sdl", geom = "errorbar") +
  stat_summary(fun.data = "mean_sdl")

met_avg %>%
  filter(!(region %in% NA)) %>%
  ggplot() +
  aes(x = region, y = dew.point) +
  stat_summary(fun.data = "mean_sdl", geom = "errorbar") +
  stat_summary(fun.data = "mean_sdl")

```

In the first plot comparing the standard deviations of the average windspeeds across the 4 regions, we see that there is a surprisingly large amount of variability in the average windspeed in all 4 regions. This is in contrast to the second plot comparing the standard deviations of the average dew point temperatures across the 4 regions, where we can see that there is high variability in the northwest and southwest but lower variability in the northeast and southeast.

### 7. Make a map showing the spatial trend in relative h in the US

-   Make sure to remove `NA`
-   Use leaflet()
-   Make a colour palette with custom colours
-   Use `addMarkers` to include the top 10 places in relative h (hint: this will be useful `rank(-rh) <= 10`)
-   Add a legend

```{r, eval = TRUE, echo = FALSE}
top10 <- met_avg %>%
  filter(!(region %in% NA)) %>%
  arrange(desc(rh)) %>%
  slice_head(n = 10)
  
rh_pal <- colorNumeric(c('blue', 'purple', 'red'), 
                       domain = met_avg$rh,
                       na.color = NA)

met_avg %>%
  filter(!(region %in% NA)) %>%
  leaflet() %>%
  addProviderTiles('OpenStreetMap') %>%
  addCircles(lat = ~lat,
             lng = ~lon,
             color = ~rh_pal(rh),
             opacity = 1,
             fillOpacity = 1,
             label = ~paste(round(rh, 2), 'rh'),
             radius = ~elev * 20) %>%
  addMarkers(lat = ~lat, 
             lng = ~lon, 
             label = ~paste(round(rh, 2), 'rh'), 
             data = top10) %>%
  addLegend('bottomleft', 
            pal = rh_pal, 
            values = met_avg$rh, 
            title = "relative humidity", 
            opacity = 1)
```

From the above map, we can see a general trend of increasing relative humidity as we move across the United States from west to east, with a few notable exceptions in the west coast states of Washington, Oregon, and California. 

### 8. Use a ggplot extension

-   Pick and extension (except cowplot) from [here](https://exts.ggplot2.tidyverse.org/gallery/) and make a plot of your choice using the met data (or met_avg)
-   Might want to try examples that come with the extension first (e.g. ggtech, gganimate, ggforce)

```{r, echo=FALSE}
ggplot(met_avg, aes(factor(region), dew.point)) + 
  geom_boxplot() + 
  # Here comes the gganimate code
  transition_states(
    round(wind.sp),
    transition_length = 2,
    state_length = 1
  ) +
  enter_fade() + 
  exit_shrink() +
  ease_aes('sine-in-out')
```

